From physical conditions to Hamiltonian and Lagrangian mechanics

### 2020-12-21 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Update classical-phil-article.tex
Wrote section 5.2 (gauge redundancy) and parts of Section 4 (significance of invariant density); implementing section outlines from last week.

First, a system is \textit{infinitesimally reducible} provided that the states of its constituent, infinitesimal parts completely specify the overall state of the system. Colloquially, we can say that the whole state is equal to the sum of its parts, where we take these infinitesimal parts to be classical particles. When this condition is satisfied, we can define a distribution $\rho$ over particle states. Furthermore, since particle states are independent of a choice of units, this distribution must be invariant. We show in Section~\ref{infinitesimal} that invariant distributions can only be defined over the phase space of classical mechanics, a symplectic manifold. The physical significance of this symplectic manifold is thereby its connection with the physical condition of infinitesimal reducibility. [To do: Clarify the scope of our treatment: is it restricted to $N$-particle classical mechanics? Or does it also handle rigid bodies and classical continual mechanics? Cite Mark Wilson 2013 on these issues]. 
% % I think I could spend more time clarifying reducibility, infinitesimal. Maybe define separately first?

@@

* "Fundamentality" because it makes the physics more prominent.

* Invariant densities: physical significance connected to coordinate independence.

* Density comparison is independent of coordinates. We can compare densities point-wise without worry.

* Contrast this with densities over position and velocity. But! We can understand some aspects of Riemmanian geometry in terms of how states densities are represented over position and velocity. Hyper-regularity allows us to express densities from "states" to "kinematic variables". The fact that this expression depends only on position (and not on velocity), tells us we have a Riemannnian manifold. Norm of the metric tensor gives us the ``conversion'' factor. Note that norm of the metric tensor in general depends on the coordinate system and is in general point-dependent. There always exists a coordinate system in which the norm of the metric tensor is constant. The conversion factor for marginal distributions is linked to the eigenvalues along each coordinate. Flatness means there exists a coordinate system where we can factorize the distribution (if it factorizes in phase space), which is exactly the Cartesian coordinate system. (Note, we can always factorize in phase space) -> Therefore this focus on densities also allows us to get a better intuition even on the tangent space.

@@

* Not all "state variable" are physical. Hamiltonian mechanics does not really care what the q's and p's are. It is the motion assumption that tells us that q's are position, simply because they define the unit system. P is a guage quantity (we care about the trajectory, the velocity). In this view, p is simply a ``label'' we can use to express the density over states. Therefore we do care about the physicality of the invariant distribution more than the physicality of the state variables themselves. -> Invariant density is really the important physical element/object in Hamiltonian mechanics.

We have now substantiated the claim of Section~\ref{privileging} that Hamiltonian mechanics requires less structure than Lagrangian mechanics, in virtue of requiring fewer physical conditions. While this provides one reason for privileging Hamiltonian mechanics, it is not the only one. A further---possibly even more important---reason stems from the relationship between Hamiltonian mechanics and invariant distributions. As we argue in this section, viewing physics through the lens of invariant distributions has much to commend it. In general, it provides a better understanding of classical systems.

Section~\ref{action} provides a further application of invariant distributions, providing a geometrical interpretation of the action principle.

Section~\ref{fundamentality} relied on a notion of fundamentality in terms of generality. Here, we consider an epistemic sense of ``fundamental." In this sense, Hamiltonian mechanics is ``fundamental" because it shows that densities are coordinate-independent quantities. Assuming that coordinate-independence is sufficient for being physically significant, this shows that these densities are physically significant. More precisely, they are physical quantities or properties of the system.

[To do: expand on the following point, and try to eliminate the notion of ``simplicity" and replace it with a more epistemic notion. Ability to make comparisons of density, independently of which coordinate system you are in or knowledge of being in a ``good" coordinate system. Josh: see 11-30-20 meeting notes where I think we discussed this.]: Compared with Lagrangian mechanics, canonical pairs (position and conjugate momentum) allow a simpler comparison of densities over space.

[To do: expand on the following point: ] This is why statistical mechanics works best on phase space (we have a uniform measure over q/p).

@@

While it is important to provide a physical interpretation of invariants, this is not to say that only invariant quantities are physically meaningful.  For instance, the coordinates q define the units, and units are physically significant despite not being invariant. Units are physically significant because they allow us to assign numbers to a physical quantity. They are physically significant in part because they correspond to measurable quantities. Whereas North claims that coordinate choices are ``merely arbitrary choices of description" \parencites*[61]{North}, some choices of the coordinates q are less physically meaningful because they are not properly connected with measurement. Some arbitrary mixings of q's and p's preserve the symplectic manifold but are not as physically significant. For this reason, the q's are epistemically privileged because they are linked with measurements, in a sense that the logically-derivative p's are not. Whereas symplectic geometry creates a mathematical, syntactic symmetry between the q and p variables, this is not a physically significant symmetry from the standpoint of measurement.

@@

@@

[Not sure if/where the following claim is relevant: ] It follows from the way in which position quantities fix the units that $\dot{x} = v $. It is built into the notion of a dynamical system that half of the state variables are derivatives of the other half.

@@

@@

In some sense (to be made more precise), gauge transformations are more prevalent in Lagrangian mechanics than Hamiltonian mechanics. This is so despite the fact that the conjugate momentum $p $ is itself a gauge quantity.

[To do: clarify the following claim:] motion equivalence allows us to understand the gauge redundancy of the Lagrangian. However, it is not true that the Lagrangian has these redundancies in virtue of the system satisfying motion equivalence.

@@

Gauge redundancy of conjugate momentum: The $p $ variables are not of fundamental significance because we can perform a gauge transformation on the $p $'s such that the $q $ and $\dot{q} $ variables remain the same. For systems where motion equivalence holds, we are really measuring the $q $ and $\dot{q} $ variables, indicating that this is a genuine gauge transformation of the momentum variables. Hence, privileging Hamiltonian mechanics does not entail, pace North, that ``momentum is a truly fundamental quantity, on a par with position" \parencites*[77]{North}. Instead, position and kinetic momentum remain physically privileged, with conjugate momentum relegated to a gauge quantity (despite its role in the Hamiltonian formalism).

@@

North assumes that formulating a Lagrangian theory requires specifying a Riemannian metric; according to Curiel, this is neither necessary nor sufficient. Furthermore, the physical significance of a Riemannian metric in the context of a Lagrangian theory is unclear: two different Lagrangian's can have the same Lagrangian vector field and yet induce inequivalent Riemannian metrics

@@

\section{Discussion of relation to categorical equivalence proofs}

>>>>

First, a system is \textit{infinitesimally reducible} provided that the states of its constituent, infinitesimal parts completely specify the overall state of the system. Colloquially, we can say that the whole state is equal to the sum of its parts, where we take these infinitesimal parts to be classical particles. When this condition is satisfied, we can define a distribution $\rho$ over particle states. Furthermore, since particle states are independent of a choice of units, this distribution must be invariant. We show in Section~\ref{infinitesimal} that to define this invariant distribution, we require a symplectic manifold, which serves as the phase space of classical mechanics. The physical significance of this symplectic manifold is thereby its connection with the physical condition of infinitesimal reducibility. [To do: Clarify the scope of our treatment: is it restricted to $N$-particle classical mechanics? Or does it also handle rigid bodies and classical continual mechanics? Cite Mark Wilson 2013 on these issues]. 
% % I think I could spend more time clarifying reducibility, infinitesimal. Maybe define separately first?

@@

* ``Fundamentality" because it makes the physics more prominent.

* Invariant densities: physical significance connected to coordinate independence. [incorporated]

* Density comparison is independent of coordinates. We can compare densities point-wise without worry. [incorporated]

* Contrast this with densities over position and velocity. But! We can understand some aspects of Riemmanian geometry in terms of how states densities are represented over position and velocity. Hyper-regularity allows us to express densities from "states" to "kinematic variables". The fact that this expression depends only on position (and not on velocity), tells us we have a Riemannnian manifold. Norm of the metric tensor gives us the ``conversion'' factor. Note that norm of the metric tensor in general depends on the coordinate system and is in general point-dependent. There always exists a coordinate system in which the norm of the metric tensor is constant. The conversion factor for marginal distributions is linked to the eigenvalues along each coordinate. Flatness means there exists a coordinate system where we can factorize the distribution (if it factorizes in phase space), which is exactly the Cartesian coordinate system. (Note, we can always factorize in phase space) -> Therefore this focus on densities also allows us to get a better intuition even on the tangent space. [Still need to incorporate much of this discussion]

@@

*[Could save the following for discussion of gauge redundancy in section 5.2] Not all ``state variables" are physical. Hamiltonian mechanics does not really care what the q's and p's are. It is the motion assumption that tells us that q's are position, simply because they define the unit system. P is a guage quantity (we care about the trajectory, the velocity). In this view, p is simply a ``label'' we can use to express the density over states. Therefore we do care about the physicality of the invariant distribution more than the physicality of the state variables themselves. -> Invariant density is really the important physical element/object in Hamiltonian mechanics.

[To do: is there a physically or conceptually important distinction between coordinate-invariance vs. coordinate-independence? (vs. co-variance?)]

We have now substantiated the claim of Section~\ref{privileging} that Hamiltonian mechanics requires less structure than Lagrangian mechanics, in virtue of requiring fewer physical conditions. While this provides one reason for privileging Hamiltonian mechanics, it is not the only one. A further---possibly even more important---reason stems from the relationship between Hamiltonian mechanics and invariant densities. As we argue in this section, viewing physics through the lens of invariant densities has much to commend it. In general, it provides a better understanding of classical systems. Specifically, we will see that invariant densities are of fundamental physical significance due to their coordinate-independence [invariance?]. Hamiltonian mechanics---unlike Lagrangian mechanics---allows us to compare these densities point-wise in any coordinate system. Although the density is physically significant because it is coordinate-invariant, we will nevertheless argue that coordinate-invariance is not a necessary condition for physical significance. In fact, the ability to express densities in some choice of units is closely connected with symplectic structure. Hence, we will argue that some coordinate-dependent representational choices are nevertheless physically significant. This view is at odds with metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms [get citations: Field 1980; Chen papers; North]. 
% Section~\ref{action} uses the invariant density to provide a geometrical interpretation of the action principle.

In the Hamiltonian formalism, the density has the following desirable property: we can compare the density point-wise across phase space. If the scalar value of the density at one point is larger than another, this entails that the density is higher at that point of phase space. Physically, this means that there is more physical stuff at this point. Likewise, if the scalar value of the density is the same at two different points, this entails that there is the same amount of stuff at those points. For brevity, will refer to this as the property of \textit{point-wise comparison}. Since the density depends in general on the choice of units, point-wise comparison does not hold universally across representations of state space. Instead, for the density of states to have this property, we require a coordinate-independent definition of the density. As the derivation in Section~\ref{deterministic} shows, we must define our state space using a symplectic manifold in order to define an invariant density. In other words, to define a density with the property of point-wise comparison, we must use canonical coordinates $(q, p)$ defined on a symplectic manifold. As we will discuss further in Section~\ref{gauge}, the conjugate momentum $p$ is itself a gauge-dependent quantity. Hence, the real physical significance of canonical coordinates is that they allow us to define an invariant phase space density $\rho $.

It is instructive to contrast this situation with Lagrangian mechanics. In this case, the density depends on the position and velocity state variables $(q,\dot{q}) $. Crucially, these state variables transform contravariantly, rather than covariantly. Consequently, the density $\rho(q,\dot{q})$ also varies contravariantly. These transformation properties preclude the density---in the Lagrangian formalism---from being coordinate-invariant in general. Of course, there exist special coordinate systems that allow for faithful point-wise comparison of the density. For instance, Cartesian coordinates---for a system in an inertial frame of reference---support this formal representational property. Yet, we still need to know which coordinate system we are in (or have transformed to) before making these point-wise comparisons. In general, non-Cartesian or non-inertial coordinate systems will not allow for point-wise comparisons of the density. The Hamiltonian formalism circumvents this further epistemic requirement by defining the density using quantities that vary covariantly. Hence, we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in.

[Should we briefly illustrate or gesture at an illustration of a failure of point-wise comparison? E.g., consider a situation where the density is a constant $k$ as a function of Cartesian coordinates (x, y). If we transform to polar coordinates $(r,\theta) $, then the scalar value of the density is no longer a constant (due to the Jacobian of this transformation not being unity). For instance, the scalar value of the density increases as we go further out, even though the density is physically the same. Hence, to determine whether there is more stuff at one location in state space vs. another, we need to know which coordinate system we are in, and how the density depends on these coordinates (i.e. we can't just compare the values of $\rho$ in these coordinates and see which is larger).]

% Connect this back with fundamentality: epistemic notion of the fundamental (don't need to know which coordinate system we are in), along with a physical sense: supporting a physical invariant, namely the invariant phase space density. Showing that this is more physically significant than the state variables themselves.

Section~\ref{fundamentality} analyzed fundamentality in terms of generality. We argued that Hamiltonian mechanics is more fundamental than Lagrangian mechanics in part because it is more general. By considering how Hamiltonian mechanics supports faithful point-wise density comparisons, we have seen two further notions of fundamentality. The first is epistemic: by representing the density on a symplectic manifold, we do not need to know which coordinate system we have chosen to work in. Any and all coordinate systems will support the formal property of point-wise comparisons [what about time-dependent coordinate transformations?]. Not only is this formal property methodologically advantageous, but it also bespeaks a further physical significance. This brings us to a second, physical notion of fundamentality: the phase space density $\rho$ is physically fundamental because it is a coordinate-invariant. Assuming that coordinate-invariance is sufficient for physical significance, the density thereby represents a genuine physical property of classical systems. It quantifies the amount of stuff in each phase space state. [to do: should we make this precise in terms of representational theory of measurement? E.g., the particular kind of scale that this representation defines and the meaningful comparisons it supports?].

% Here, we consider an epistemic sense of ``fundamental." In this sense, Hamiltonian mechanics is ``fundamental" because it shows that densities are coordinate-independent quantities. Assuming that coordinate-independence is sufficient for being physically significant, this shows that these densities are physically significant. More precisely, they are physical quantities or properties of the system.

[To do: expand on the following point: ] This is why statistical mechanics works best on phase space (we have a uniform measure over q/p). See the relevant * point above.

@@

While it is important to provide a physical interpretation of invariants, this is not to say that only invariant quantities are physically meaningful.  For instance, the coordinates $q$ define the units, and units are physically significant despite not being invariant. Units are physically significant because they allow us to assign numbers to a physical quantity. They are physically significant in part because they correspond to measurable quantities. Indeed, as the derivation in Section~\ref{deterministic} shows, a necessary condition for defining the symplectic structure is that we can express the density in a particular choice of units. If we did not have the physical ability to define the density in some units---although any particular choice is arbitrary---we would not be able to define this structure. [To do: make this argument more clear]

Whereas North claims that coordinate choices are ``merely arbitrary choices of description" \parencites*[61]{North}, some choices of the coordinates $q$ are less physically meaningful because they are not properly connected with measurement. Some arbitrary mixings of $q$'s and $p$'s preserve the symplectic manifold but are not as physically significant. For this reason, the $q$'s are epistemically privileged because they are linked with measurements, in a sense that the logically-derivative $p$'s are not. Whereas symplectic geometry creates a mathematical, syntactic symmetry between the $q$ and $p$ variables, this is not a physically significant symmetry from the standpoint of measurement.
%cite van Fraassen on the measurement point; e.g. APA session on Halvorson book

@@

% Have mostly done the following: [To do: expand on the following point, and try to eliminate the notion of ``simplicity" and replace it with a more epistemic notion. Ability to make comparisons of density, independently of which coordinate system you are in or knowledge of being in a ``good" coordinate system. Josh: see 11-30-20 meeting notes where I think we discussed this.]: Compared with Lagrangian mechanics, canonical pairs (position and conjugate momentum) allow a simpler comparison of densities over space.

@@

% By definition, a dynamical system is one where we can write xdot as a function of x (i.e. the current state determines the evolution of the system). Having a Hamiltonian system further means that we can express this function xdot(x) using the Hamiltonian.

@@

% 12-19: I think that I incorporated the relevant points below:

@@

Understanding the action principle in terms of Hamiltonian mechanics also provides a better understanding of gauge transformations. If motion equivalence holds, then gauge transformations leave the variables $(q,\dot{q}) $ invariant. To see this, recall that motion equivalence entails that knowledge of the trajectory in physical space suffices for knowledge of the state in phase space. In other words, we can completely characterize states in terms of the canonical position coordinate $q$ and its first derivative $\dot{q}$. Hence, any transformation that leaves $(q,\dot{q}) $ invariant \textit{ipso facto} leaves the physical state invariant, i.e. is a gauge transformation. [To do: make it more clear how the discussion in this section is intrinsically connected with the earlier geometric treatment of the action and our overall privileging of Hamiltonian mechanics]

In particular, there is a family of gauge transformations that changes the conjugate momentum variable $p$ while leaving $(q,\dot{q}) $ invariant. [To do: illustrate this with general formula for this kind of gauge transformation, and possibly the electrodynamics example.] Consequently, the conjugate momentum in Hamiltonian mechanics is gauge-dependent. The $p $ variables are therefore not of fundamental physical significance [to do: check that this is consistent with Section 2.1's discussion of fundamentality]. Hence, privileging Hamiltonian mechanics does not entail, \textit{pace} North, that ``momentum is a truly fundamental quantity, on a par with position" \parencites*[77]{North}. [To do: see if Wallace 2003 also makes this claim about the fundamentality of conjugate momentum, or the similarly flawed claim that conjugate momentum and position are on equal footing within Hamiltonian mechanics]. Although the canonical position and momentum variables are on equal syntactical footing within symplectic mechanics, this does not entail that they are equally physically significant.

Instead, position and kinetic momentum are physically privileged, since these quantities are gauge-invariant. Consequently, Lagrangian mechanics employs state space variables that are gauge-invariant, while Hamiltonian mechanics uses a gauge-dependent variable. Despite this advantage, Lagrangian mechanics has a compensating disadvantage: the Lagrangian is itself gauge-dependent. The formalism of Section~\ref{action} makes this gauge dependence manifest. Representing the Lagrangian as $L =\vector{\theta} \cdot \vector{S} $, we see that since the vector potential $\vector{\theta}$ is gauge-dependent while the displacement vector $\vector{S}$ is not, $L$ must also be gauge-dependent.\footnote{As shown in Section~\ref{action} we can arbitrarily transform $\vector{\theta}$ by adding the gradient of an arbitrary scalar function, while $\vector{S}=curl(-\vector{\theta}) $ remains invariant under this gauge transformation.} The gauge-dependence of the Lagrangian makes it difficult to provide a clear physical interpretation of any particular Lagrangian.\footnote{The action inherits the gauge-dependence of the Lagrangian. This does not affect the principle of stationary action because the local minima and maxima of the action remain invariant under gauge transformations.} In contrast, the Hamiltonian is defined up to a constant, endowing it with greater physical significance. Overall then, we see that neither Lagrangian nor Hamiltonian mechanics has the upper hand when it comes to invoking only gauge-independent physical quantities. Both formalisms have their trade-offs in this regard (although, it is worth emphasizing that---as argued in Section~\ref{density}---Hamiltonian mechanics provides the natural setting for the invariant phase space density).

Moving from gauge transformations to coordinate transformations, we gain further reasons for using Hamiltonian mechanics within the extended phase space formalism. For many applications, it suffices to use time-independent coordinate transformations. In this special case, both the Lagrangian $L$ and the Hamiltonian $H$ are coordinate-invariant. However, some applications benefit from (or even require) the use of time-dependent coordinate transformations, e.g. a rotating frame of reference. In this case, the Lagrangian remains coordinate-invariant, while the Hamiltonian becomes coordinate-dependent. Again, the formalism of Section~\ref{action} makes this time-dependence manifest. As we saw there, the Hamiltonian $H$ is the time component of the vector potential $\vector{\theta} $, entailing that $H$ transforms as the time component of a co-vector. Methodologically, this makes the Hamiltonian framework less convenient for treating this class of problems.

Nonetheless, formulating Hamiltonian mechanics within the extended phase space provides one way to rectify these deficiencies. At root, the time-dependence of the Hamiltonian stems from the dual role of time as both a coordinate and an affine parameter within the usual Hamiltonian framework. Yet, differential geometry demands that we separate these two roles. We thereby add a temporal degree of freedom $(t, E) $, expanding the phase space degrees of freedom by one (recall that we previously had one degree of freedom for every canonical coordinate pair $q_i, p_i$). Next, we define the phase space evolution over an affine parameter $s $. In this extended Hamiltonian framework, the Hamiltonian constraint functions as both a global constraint---fixing the energy as a function of $q, p, t $---and as a generator of the affine parameter $s $ [to do: spell out what the `Hamiltonian constraint' is in general]. Crucially, the Hamiltonian constraint is now coordinate-invariant, even with respect to time-dependent coordinate transformations. For example, if we define the affine parameter as the proper time, then the Hamiltonian constraint is proportional to the mass (which is manifestly coordinate-invariant). This formalism thereby provides a clear physical meaning for the Hamiltonian constraint. [To do: provide some reference for this modified Hamiltonian formulation]

[To do: not sure how best to end this subsection/transition to the next one]. We have now considered many conceptual and physical advantages of the Hamiltonian formalism. In Section~\ref{Curiel}, we turn to considering Curiel's \parencites*[]{Curiel} arguments against privileging Hamiltonian mechanics.

[Could address the following in this section, rather than in Section 4:] Not all ``state variables" are physical. Hamiltonian mechanics does not really care what the q's and p's are. It is the motion assumption that tells us that q's are position, simply because they define the unit system. P is a gauge quantity (we care about the trajectory, the velocity). In this view, p is simply a ``label'' we can use to express the density over states. Therefore we do care about the physicality of the invariant distribution more than the physicality of the state variables themselves. This entails that the invariant density is really the important physical element/object in Hamiltonian mechanics.

I think that I have incorporated most of the following into the above:

I didn't incorporate the following point: In some sense (to be made more precise), gauge transformations are more prevalent in Lagrangian mechanics than Hamiltonian mechanics. This is so despite the fact that the conjugate momentum $p $ is itself a gauge quantity.

% [To do: clarify the following claim:] motion equivalence allows us to understand the gauge redundancy of the Lagrangian. However, it is not true that the Lagrangian has these redundancies in virtue of the system satisfying motion equivalence.

@@

@@

North assumes that formulating a Lagrangian theory requires specifying a Riemannian metric; according to Curiel, this is neither necessary nor sufficient. Furthermore, the physical significance of a Riemannian metric in the context of a Lagrangian theory is unclear: two different Lagrangians can have the same Lagrangian vector field and yet induce inequivalent Riemannian metrics

@@

\section{Material looking for a possible home}
